Co-citations in context

### 2019-06-10 cocit_r.tex

%%%%%%%%%%% to be supplied by MIT Press, only %%%%%%%%%%%%%%%%%
\citation{Betzel, R. F., Fukushima, M.,
He, Ye,
Zuo, Xi-Nian,
Sporns, O. (2016)\\
Dynamic fluctuations coincide with periods of high and low modularity
in resting-state functional brain  networks\\
Network Neuroscience, 1
}

@@

\citep{uzzi_atypical_2013} used a novel approach for co-citation analysis using, as input, 17.9 million articles and their cited references, from the Web of Science (WoS)  to characterize a subset of highly cited articles with respect to both novel and conventional combinations of prior research. In this study, co-citation frequencies of references in were calculated and represented as journal pairs (observed co-citation frequencies). Expected values of these frequencies were generated from Monte Carlo simulations under a random graph model. Observed frequencies were then normalized (shifted and scaled) to averaged expected values from ten simulations and termed \emph{z-scores}, which were interpreted as suggesting conventional thought if positive and novel thought if negative. Consequently, every article was associated with multiple z-scores corresponding to co-cited journal pairs in its refernces. For each article, positional statistics of z-scores were calculated to set thresholds for a binary (high/low) classification of conventionality (C) using the median and novelty (N) using the tenth percentile of z-scores within an article. Thus, HCLN would denote high conventionality and low novelty with all four combinations being possible. 
\cite{uzzi_atypical_2013} observed that HCHN articles were twice as likely to be highly cited, suggesting that novel combinations of ideas flavoring a body of conventional thought were a 'nearly universal` recipe for impact.

Key to Uzzi {\em et al.}, is their random graph model and its underlying assumptions. The citation switching algorithm used to generate a null model elegantly preserves the number of publications, the number of references in each publication, the year of publication of both publications and references, and the disciplinary composition of the references cited in these publications. However, random substitution of the references used to generate expected values is equiprobable with respect to disciplinary origin and citation count. The model permits substitution of a reference in quantum physics with equal probability by a reference in the related quantum chemistry field or in some entirely different field, such as classical literature, evolutionary biology, or anthropology. Such substitutions poorly model the disciplinary nature of scientific endeavor and citation behavior~\citep{wallace_lariviere_gingras_2012,moed_measuring_2010,klavans_research_2017,garfield_1979}. In addition, under this random model, a reference cited over 100 times in a given year is selected with the same probability as a reference cited only once, which appears inconsistent with the power law or lognormal citation distributions described in the literature~\citep{stringer_statistical_2010,perline_strong_2005}.  Accordingly, the expected value calculations used to characterize articles in terms of conventionality and novelty can be reasonably questioned on grounds of model mis-specification when disciplinary subsets of the WoS are analyzed in the context of expected values generated across all of WoS.

@@

Despite different methods used to generate expected values, both of these key preceding studies (vide supra) measured co-citation frequencies across the scientific literature without disciplinary constraints and subsequently analyzed disciplinary subsets. In extending their prior work, we hypothesized that normalization using expected values drawn from datasets focused in disciplinary areas, rather than all of WoS, would reduce model mis-specification.  Accordingly, we used keyword searches of the scientific literature to create three citation networks themed around academic disciplines. Within these disciplinary frameworks, we calculated observed and expected co-citation frequencies using a refined random graph model and an efficient Monte Carlo simulation algorithm. Our analysis of these semantically-themed citation networks challenges the assertion of universality. Under our conditions of reduced model misspecification, we observe that while HC remains highly correlated with hit articles in the immunology and metabolism datasets, HN is not. In addition, while HN is highly correlated with hit articles in applied physics, HC is not. Beyond variation across disciplines and the case for contextual analysis, we found that the categories demonstrating the highest percentage of hits are not robust with respect to varying parameters that define hit articles and the threshold for highly novel citation patterns.

@@

\begin{table}[ht]
\caption{Disciplinary Datasets. PubMed and WoS were searched for articles using search terms, `immunology', `metabolism', and `applied physics'. Counts of retrieved publications are shown for each of the three years analyzed.}
\label{tab:disc-datasets}
\centering
\begin{tabular}{|r r r r|}
  \hline
Year & Immunology & Metabolism & Applied Physics \\ 
  \hline
1985 & 21,606 & 78,998 & 10,298 \\ 
1995 & 29,320 & 121,247 & 21,012  \\ 
2005 & 37,296 & 200,052 & 35,600  \\ 
 \hline
\end{tabular}
\end{table}

\emph{Monte Carlo simulations, normalization of observed frequencies, annotations, and `hit' papers}. Analysis was performed on publications from 1985, 1995, and 2005. Building upon prior work~\cite{uzzi_atypical_2013}, all ${n \choose 2}$ reference pairs were generated for each publication, where $n$ is the number of cited references in the publication. These reference pairs were then mapped to the journals they were published in using ISSN numbers as identifiers. Where multiple ISSN numbers exist for a journal, the most frequently used one in the WoS was assigned to the journal. In addition, publications containing fewer than two references were discarded. Journal pair frequencies were summed across the dataset to create observed frequencies $(F_{obs})$. We developed a performant citation switching algorithm, \emph{runtime enhanced permuting citation switcher (repcs)} that randomly permuted citations grouped by year of publication to switch citations while preserving the year of publication for both articles and references, the number of publications and the number of references in each dataset and the disciplinary composition of the references in each dataset. We used the citation switching algorithm of ~\cite{uzzi_atypical_2013} referred to as \emph{umsj}~\citep{boyack_vs_uzzi_2014}, as a benchmark for the repcs algorithm using code kindly provided by the authors. A comparative analysis showed that while 10 simulations of the WoS 1985 dataset (391,860 publications) completed in 2,186 hours using the umsj algorithm, it completed in less than one hour using the repcs algorithm implemented on a Spark cluster. Using either the repcs or umsj algorithms for 10 simulations resulted in expected value coverage of 75\% of the observed journal pair frequencies. Subsequent rounds of development and testing resulted in 99\% coverage of observed journal pairs when 1,000 simulations were conducted. The runtime advantage was significant enough that we chose to use the repcs algorithm in our study  and generated expected values averaged from 1,000 simulations for improved coverage for every dataset we analyzed.

@@

We  next analyzed the consequences of citation shuffling within a disciplinary set or all of the Web of Science.  References in publications belonging to these three datasets were classified using the WoS Extended Subjects classification of 153 subject areas and summarized as a frequency distribution. A single shuffle of the references in these datasets or all the references in the corresponding WoS year slice was performed, using either the rpcs or umsj algorithms, after which subject frequencies were computed again. The fold change in subject frequencies of references before and after shuffling was calculated for these groups using all 153 Subject categories and are summarized as boxplots in Fig \ref{fig:be}. For example, the applied physics dataset contained 1 reference labeled Genetics and Heredity . This proportion was preserved in the intra-network shuffle but increased 1496 times when the WoS background was used as a source of substitution during Monte Carlo simulation with the repcs algorithm. Similarly, with the metabolism dataset.which contained one reference labeled Philosophy that was increased to 661 after a single repcs shuffle. The data show convincingly that the disciplinary composition of references is preserved when citation shuffling is conducted within a disciplinary network but is significantly altered when the full WoS network is used as a source of substitution. A second inference is that the choice of algorithm is not relevant beyond runtime advantage but that the background network used for the shuffle is critical.

@@

\begin{figure}
\includegraphics[width=\hsize]{Fig1}
\caption{Fraction of publications with high conventionality (HC) and high novelty (HN) signatures relative to citation count.      \emph{What follows is the caption from PNAS submission: is any of this useful?}  Effect of Research Discipline,  Background Network,  and Citation Count on Conventionality and Novelty.
Data are shown for the applied physics (18,305), immunology (21,917), metabolism (97,405) and WoS (476,288) networks for 1995. The number of publications in each network is shown in parentheses. Citation counts shown are cumulative over the first 8 years since publication. 
X-axis: publications were classified into percentile groups based on citation counts (e.g., Top 1 indicates those publications in the top 1\%). Y axis: The percent of applications in each group that are high conventionality (HC) and high novelty (HN). The z-scores are computed for each disciplinary network based on the selected background network; thus, \emph{imm} denotes the immunology network with immunology z-scores and \emph{imm\_wos} denotes the immunology network with z-scores from WoS z-scores. The figure shows striking differences between the WOS network compared to the metabolism and immunology networks: across all networks, the percentage of high conventionality (HC) publications increases with citation counts, while the percentage of high novelty (HN) publications decreases with citation counts for the biological networks but not for WOS.
\label{fig:Fig1}}
\end{figure}

Another symptom of model misspecification is that random sampling of citations associated with an article do not match with the distribution of citations across the disciplines observed in articles from that discipline. \citep{wallace_lariviere_gingras_2012} found that, often, a majority of an article's citations are from the specialty of the article, although that percentage varied among disciplines in the eight specialties they investigated: from approximately 39\% to 89\% for 2006. \emph{More needed here.}  Similarly, we found that 77.4\% of citations in applied physics articles in 1985 (replace with 1995 data?) were from applied physics.  \emph{Add references to other datasets.} A random sampling of citations would ideally reflect the percentages from the discipline in question, as well as the percentages of articles from other disciplines.

We investigated the effect of sampling from disciplinary data sets versus more broadly from the WoS on the disciplinary make-up of the citations in articles under random sampling (\emph{sentence needs work but is an example of a transition.}) \emph{Insert text about Figure \ref{fold_effect}, specifically the right-hand side of Figure  \ref{fold_effect} where the fold statistics demonstrate the effect with Uzzi's approach.}

\emph{Insert comment about the left-hand side of Figure \ref{fold_effect} where the disciplinary dataset has resolved the fold issue.}  We also predicted that model misspecification as measured by the Kullback-Leibler (K-L) Divergence~\citep{kullback_information_1951} between observed and simulated frequencies in a disciplinary network would be less than the divergence for observed and simulated frequencies in the WoS superset. The results in Table \ref{tab:kld} indicate that for the set of journals common to both a disciplinary network and the WoS superset, simulations under our model consistently have a lower K-L divergence compared to simulations that draw from the WoS superset (and its attendant substitutions that are ectopic with respect to field and discipline).

@@

We conclude that Uzzi et al.'s finding of high hit rates in the HNHC category does not hold generally for all datasets and that novel citation patterns are not always indicative of impactful research.

The z-scores that change sign across datasets is either contradictory or an acceptable variation due to the difference in reference sets. We contend that it is the former because the inappropriate substitution of citations with those from disciplines that are implausible and because the observed citation frequencies are ignored. That fewer z-scores are negative in the WoS dataset relative to the immunology dataset may be directly due to the uniform sampling of references whereby many resulting journal pairs are never observed in the literature and so that the expected frequencies of those observed pairs have lower expected values, thus biasing their z-scores upward. In addition, treating citation as a set versus a multiset means that a random model will sample popular citations downward, thus further increasing their z-scores.

@@

George's point from Slack conversation: ``A concern that surfaced on the drive are the definitions of conventionality and novelty. On the one hand we use them- on the other hand you've also pointed out that it isn't that simple.''

>>>>

%%%%%%%%%%% to be supplied by MIT Press, only %%%%%%%%%%%%%%%%%
\citation{Betzel, R. F., Fukushima, M.,w
He, Ye,
Zuo, Xi-Nian,
Sporns, O. (2016)\\
Dynamic fluctuations coincide with periods of high and low modularity
in resting-state functional brain  networks\\
Network Neuroscience, 1
}

@@

\citep{uzzi_atypical_2013} used a novel approach for co-citation analysis using, as input, 17.9 million articles and their cited references, from the Web of Science (WoS)  to characterize a subset of highly cited articles with respect to both novel and conventional combinations of prior research. In this study, the frequency with which references were calculated and reduced to journal pairs frequencies (observed co-citation frequencies). Expected values were generated from Monte Carlo simulations under a random graph model. Observed frequencies were then normalized (shifted and scaled) to averaged expected values from ten simulations and termed as \emph{z-scores}, which were interpreted as indicating conventional thought if positive and novel thought if negative. Consequently, every article was associated with multiple z-scores corresponding to co-cited journal pairs in its refernces. For each article, positional statistics of z-scores were calculated to set thresholds for a binary (high/low) classification of conventionality (C) using the median and novelty (N) using the tenth percentile of z-scores within an article. Thus, HCLN would denote high conventionality and low novelty with all four combinations being possible. 
\cite{uzzi_atypical_2013} observed that HCHN articles were twice as likely to be highly cited, suggesting that novel combinations of ideas flavoring a body of conventional thought were a 'nearly universal` feature for impact.

Key to Uzzi {\em et al.}, is their random graph model and its underlying assumptions. The citation switching algorithm used to generate a null model elegantly preserves the number of publications, the number of references in each publication, the year of publication of both publications and references, and the disciplinary composition of the references cited in these publications. However, random substitution of the references used to generate expected values is equiprobable with respect to disciplinary origin and citation count. The model permits substitution of a reference in quantum physics with equal probability by a reference in the related quantum chemistry field or in some entirely different field, such as classical literature, evolutionary biology, or anthropology. Such substitutions poorly model the disciplinary nature of scientific endeavor and citation behavior~\citep{wallace_lariviere_gingras_2012,moed_measuring_2010,klavans_research_2017,garfield_1979}. In addition, under this random model, a reference cited over 100 times in a given year is selected with the same probability as a reference cited only once, which appears inconsistent with the power law or lognormal citation distributions described in the literature~\citep{stringer_statistical_2010,perline_strong_2005}.  Accordingly, model mis-specification is likely to arise from the expected value calculations used to generate z-scores and characterize articles in terms of conventionality and novelty when disciplinary subsets of the WoS are analyzed in the context of expected values generated across all of WoS.

@@

Despite different methods used to generate expected values, both of these key preceding studies (\emph{vide supra}) measured co-citation frequencies across the scientific literature without disciplinary constraints and subsequently analyzed disciplinary subsets. In extending their prior work, we hypothesized that normalization using expected values drawn from datasets focused in disciplinary areas, rather than all of WoS, would reduce model mis-specification.  Consequently, we used keyword searches of the scientific literature to construct citation networks themed around academic disciplines: \emph{applied physics, immunology, and metabolism}. Within these disciplinary frameworks, we calculated observed and expected co-citation frequencies using a refined random graph model and an efficient Monte Carlo simulation algorithm.

Our analysis of these semantically-themed citation networks challenges the assertion of universality. We demonstrate reduced model misspecification and observe that while HC remains highly correlated with hit articles in the immunology and metabolism datasets, HN is not. In addition, while HN is highly correlated with hit articles in applied physics, HC is not. Thus, disciplinary networks are different from each other and the full WoS network. Beyond variation across disciplines and the case for contextual analysis, we found that the categories demonstrating the highest percentage of hits are not robust with respect to varying parameters that define hit articles and the threshold for highly novel citation patterns.

@@

\begin{table}[ht]
\caption{Disciplinary Datasets. PubMed and WoS were searched for articles using search terms, `immunology', `metabolism', and `applied physics'. Counts of publications are shown for each of the three years analyzed.}
\label{tab:disc-datasets}
\centering
\begin{tabular}{|r r r r|}
  \hline
Year & Immunology & Metabolism & Applied Physics \\ 
  \hline
1985 & 21,606 & 78,998 & 10,298 \\ 
1995 & 29,320 & 121,247 & 21,012  \\ 
2005 & 37,296 & 200,052 & 35,600  \\ 
 \hline
\end{tabular}
\end{table}

\emph{Monte Carlo simulations, normalization of observed frequencies, annotations, and `hit' papers}. Analysis was performed on publications from 1985, 1995, and 2005. Building upon prior work~\cite{uzzi_atypical_2013}, all ${n \choose 2}$ reference pairs were generated for each publication, where $n$ is the number of cited references in the publication. These reference pairs were then mapped to the journals they were published in using ISSN numbers as identifiers. Where multiple ISSN numbers exist for a journal, the most frequently used one in the WoS was assigned to the journal. In addition, publications containing fewer than two references were discarded. Journal pair frequencies were summed across the dataset to create observed frequencies $(F_{obs})$.

We developed a performant citation switching algorithm, \emph{runtime enhanced permuting citation switcher (repcs)} that randomly permuted citations grouped by year of publication to switch citations while preserving the year of publication for both articles and references, the number of publications and the number of references in each dataset and the disciplinary composition of the references in each dataset. We used the citation switching algorithm of ~\cite{uzzi_atypical_2013} referred to as \emph{umsj}~\citep{boyack_vs_uzzi_2014}, as a benchmark for the repcs algorithm using code kindly provided by the authors. A comparative analysis showed that while 10 simulations of the WoS 1985 dataset (391,860 publications) completed in 2,186 hours using the umsj algorithm, it completed in less than one hour using the repcs algorithm implemented on a Spark cluster. Using either the repcs or umsj algorithms for 10 simulations resulted in expected value coverage of 75\% of the observed journal pair frequencies. Subsequent rounds of development and testing resulted in 99\% coverage of observed journal pairs when 1,000 simulations were conducted. The runtime advantage was significant enough that we chose to use the repcs algorithm in our study  and generated expected values averaged from 1,000 simulations for improved coverage for every dataset we analyzed.

@@

A symptom of model misspecification is that random sampling of citations associated with an article do not match with the distribution of citations across the disciplines observed in articles from that discipline. \cite{wallace_lariviere_gingras_2012} found that, often, a majority of an article's citations are from the specialty of the article, although that percentage varied among disciplines in the eight specialties they investigated: from approximately 39\% to 89\% for 2006. We also analyzed the consequences of citation shuffling within a disciplinary set or all of the Web of Science.  References in publications belonging to these three datasets were classified using the WoS Extended Subjects classification of 153 subject areas and summarized as a frequency distribution. A single shuffle of the references in these datasets or all the references in the corresponding WoS year slice was performed, using either the rpcs or umsj algorithms, after which subject frequencies were computed again. The fold change in subject frequencies of references before and after shuffling was calculated for these groups using all 153 Subject categories and are summarized as boxplots in Fig \ref{fig:be}. For example, the applied physics dataset contained 1 reference labeled Genetics and Heredity . This proportion was preserved in the intra-network shuffle but increased 1496 times when the WoS background was used as a source of substitution during Monte Carlo simulation with the repcs algorithm. Similarly, with the metabolism dataset.which contained one reference labeled Philosophy that was increased to 661 after a single repcs shuffle. The data show convincingly that the disciplinary composition of references in a network is preserved when citation shuffling is conducted within it but is significantly altered when the full WoS network is used as a source of substitution. A second inference is that the choice of algorithm is not relevant beyond runtime advantage but that the background network used for the shuffle is critical.

@@

\begin{figure}
\includegraphics[width=\hsize]{pnas_fig1a}
\caption{Effect of Research Discipline,  Background Network,  and Citation Count on Conventionality and Novelty.
Data are shown for the applied physics (18,305), immunology (21,917), metabolism (97,405) and WoS (476,288) networks for 1995. The number of publications in each network is shown in parentheses. Citation counts shown are cumulative over the first 8 years since publication. 
X-axis: publications were classified into percentile groups based on citation counts (e.g., Top 1 indicates those publications in the top 1\%). Y axis: The percent of applications in each group that are high conventionality (HC) and high novelty (HN). The z-scores are computed for each disciplinary network based on the selected background network; thus, \emph{imm} denotes the immunology network with immunology z-scores and \emph{imm\_wos} denotes the immunology network with z-scores from WoS z-scores. The figure shows striking differences between the WOS network compared to the metabolism and immunology networks: across all networks, the percentage of high conventionality (HC) publications increases with citation counts, while the percentage of high novelty (HN) publications decreases with citation counts for the biological networks but not for WOS.
\label{fig:Fig1}}
\end{figure}

@@

We conclude thathigh hit rates in the HNHC category does not hold generally for all datasets and that conventionality flavored with novelty is not always indicative of impactful research. The z-scores that change sign across datasets is either contradictory or an acceptable variation due to the difference in reference sets. We contend that it is the former because the inappropriate substitution of citations with those from disciplines that are implausible and because the observed citation frequencies are ignored. That fewer z-scores are negative in the WoS dataset relative to the immunology dataset may be directly due to the uniform sampling of references whereby many resulting journal pairs are never observed in the literature and so that the expected frequencies of those observed pairs have lower expected values, thus biasing their z-scores upward. In addition, treating citation as a set versus a multiset means that a random model will sample popular citations downward, thus further increasing their z-scores.\emph{need confirmation from Jim here}.

@@
